\section*{Acknowledgments}

This effort would have never been possible without the support and encouragement of many others.

Firstly, I would like to thank my adviser and the chair of the thesis committee, Dr. Sridhar Radhakrishnan. He was the first point of contact when I initially became interested in the Master's program at OU and had me excited about the potential from the very start. I would also like to thank the remaining members of the committee, Dr. Christan Grant and Dr. Le Gruenwald, for their role in this and for the courses I took from them.

Secondly, I would like to thank the members of the C++ and Rust communities on Reddit. Of the languages used in this research, these were the least-familiar to me and the ones I sometimes struggled with. Without the ready answers to my many questions, I would have had a much more difficult time getting the final results that I did. I would like to especially thank Andrew Gallant of the Rust community for his extensive advice and help with Rust in general and several of the programs in particular.

Next, I would like to thank my mother, whose words of support and encouragement often came at truly opportune times.

I would next like to thank my father, who passed away before I could begin this program but who would have been exceedingly proud of what I have been able to accomplish. Without his support, I would have barely managed to finish my undergraduate years at OU, let alone been in a position to return for a higher degree later in life.

Lastly, and most importantly, I thank my dear wife Eugenie. I truly believe that, without her help and support, I would not have gotten this far. She was my steady support, my cheerleader, my proofreader, my copy-editor, my sounding board for ideas, my rubber duck, and my project manager. But most of all, she was my partner. She kept me focused and kept me moving forward, even in the times I sat at our dining table and doubted my ability to finish this effort.

She never lost faith or confidence in me, and for that (among many many reasons) I dedicate this work to her.
